---
title: "MAR Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MAR Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    eval = FALSE
)
```

# The standard `mar` workflow

## Overview

This vignette demonstrates the stepwise workflow for conducting a Mutation-Area
Relationship (MAR) analysis using the `mar` package, illustrated with the included
 *Arabidopsis thaliana* demo data.

```{r setup}
library(mar)
```

## Step 1: Import raw data

Begin by locating and importing the bundled example data included with the
`mar` package. `system.file()` resolves the path to files stored in
`inst/extdata/`, ensuring the vignette works regardless of where the package
is installed. `lonlat_parser()` reads the coordinate file (TXT, CSV, or TSV) and
returns a data frame of sample IDs with their longitude and latitude, while
`text_parser()` reads the genotype matrix and returns a `margeno` object
containing the genotype data alongside sample and variant metadata.

TODO: Add marApp instructions when needed. 

```{r import}
# Sample coordinates
lonlatfile <- system.file("extdata", "1001g_lonlat.txt", package = "mar")

# Raw genotype matrix
# Tips: text_parser reads both gzipped and uncompressed text files.
genofile <- system.file("extdata", "1001g_genotypes.txt.gz", package = "mar")
# Optional: text_parser also reads sample names and chromosome positions if provided
samplefile <- system.file("extdata", "1001g_accessions.txt", package = "mar")
chrposfile <- system.file("extdata", "1001g_chrpos.txt", package = "mar")
# Tips: for your own data, change the variables above to the corresponding filepaths

# Read coordinates
# Tips: lonlat_parser will assume the coordinates are longitude and latitude following ESPG: 4326 (WGS84)
# and automatically determine the map resolution following equation 12 in (Hengl 2006)
# Optional: for alternative resolution and map projection, see Appendix S1: customize projection and map resolution
mapsdata <- mar::lonlat_parser(lonlatfile)
mapsdata <- mar::marmaps(mapsdata, mapres = NULL, mapcrs = "EPSG:8857") #should be able to get read of this line now!
print(mapsdata)
# Optional: to inspect the reference map created
plot(mapsdata)

# Read genotype
# Optional: for vcf parsing, see Appendix S2: reading vcf files
genodata <- mar::text_parser(
    geno.fn = genofile, 
    samp.fn = samplefile, 
    pos.fn = chrposfile, 
    ploidy = 2
)

print(genodata)
```

## Step 2: Construct `genomaps` object

`genomaps()` merges the `margeno` and `marmaps` objects into a single analysis
object used by all downstream functions.

```{r genomaps}
gm1001g <- genomaps(genodata, mapsdata)
print(gm1001g)
```


## Alternative Step 1-2: Load example data 

The `mar` package also provides an example dataset already in the `genomaps` data format.
If you do not need to load your own data, skip Steps1-2, and do the following:

```{r example}
data(gm1001g)
print(gm1001g)

# gm1001g is a list of two objects, to obtain marmaps and margeno objects
genodata = gm1001g$geno
mapsdata = gm1001g$maps

# Tips: most operations in `mar` package takes `genomaps` as the input. 
# Therefore, you do not have to extract the marmaps and margeno objects most if not all the time.
```



## Step 3: Explore the Site Frequency Spectrum (SFS)

The observed SFS summarises how allele counts are distributed across sites;
the expected SFS provides a neutral-theory baseline for comparison.

```{r sfs}
# Generate observed SFS from the genomaps object
obs_sfs <- sfs(gm1001g)

# Plot the observed SFS
plot(obs_sfs)

# Expected (neutral) SFS from the genomaps object
exp_sfs <- expsfs(gm1001g)
plot(exp_sfs)

# Tips: you can also generate an unfolded SFS
obs_sfs_unfolded <- sfs(gm1001g, folded = FALSE)
plot(obs_sfs_unfolded)
exp_sfs_unfolded <- expsfs(gm1001g, folded = FALSE)
plot(exp_sfs_unfolded)
```

## Step 4: Estimate the mutations-area relationship

The MAR is estimated in two steps.

### 4a. Generate the diversity-by-area sampling table

TODO: For alternative sampling schemes and other MARsampling options. See Appendix S3: Alternative MARsampling

TODO: Description here on the output table. and the default options (e.g. nrep, xfrac). Use the marApp documentation when needed.

TODO: I cannot get the plot method to register for marsamp (crying in my heart >_<), could you check this? 



```{r mar-sampling}
# uses of myseed not neccessary, used here for reproducibility
mardf <- MARsampling(gm1001g, myseed = 91)
mardf
# mardf is a marsamp object
class(mardf)
# plot the diversity-by-area sampling results (?mar::plot.marsamp)
plot.marsamp(mardf)
plot(mardf) # doesn't work but should work
```

### 4b. Fit the mutations-area relationship model

TODO: Description here

```{r mar-fit}
marfit <- MARcalc(mardf)
# marfit is a sars object (from R package sars)
class(marfit)
# To extract full mar fit table, use summary
summary(marfit)
# Extract the c and z values from marfit
c = unname(marfit$par['c'])
z = unname(marfit$par['z'])
# Plot the fit again
plot.marsamp(mardf, c = c, z = z)
```

You can also fit other mutation relationships. For example, nucleotide diversity thetapi

```{r}
# plot alternative diversity-by-area sampling results
plot(mardf, Mtype = 'thetapi', Atype = 'Asq')
marfit_pi <- MARcalc(mardf, Mtype = 'thetapi', Atype = 'Asq')

# marfit is a sars object (from R package sars)
class(marfit_pi)
# To extract full mar fit table, use summary
summary(marfit_pi)
# Extract the c and z values from marfit_pi
c = unname(marfit_pi$par['c'])
z = unname(marfit_pi$par['z'])
# Plot the fit again
plot.marsamp(mardf, c = c, z = z, Mtype = 'thetapi', Atype = 'Asq')
```

For pipelining 

## Step 5: Estimate Extinction / Diversity Loss

`MARextinction()` projects the expected loss of genetic diversity under
hypothetical area reduction scenarios.

TODO: For alternative sampling schemes and other MARextinction options. See Appendix S4: Alternative MARextinction


```{r mar-extinction}
marext <- MARextinction(gm1001g, myseed = 91)
marext

marextfit <- MARcalc(marext)
summary(marextfit)

# plot the extinction mar fit
mar:::plot.marextinct(marext, z = 3.7916e-01)
```

## Step 7: Inspect and Export Outputs

Use `str()`, `print()`, and `summary()` to inspect any result object.

```{r inspect}
# Object structures
str(mar_sampling)
str(mut_div)
str(mar_fit)
str(mar_ext)

# Model summaries
print(mar_fit)
summary(mar_fit)
```

To save results or export to CSV:

```{r export}
saveRDS(mar_fit, file = "mar_fit.rds")
write.csv(as.data.frame(mar_sampling), file = "mar_sampling.csv", row.names = FALSE)
```


# Appendix

## Appendix S1: Customizing projection and map resolution


## Appendix S2: Reading VCF files


## Appendix S3: Alternative MARsampling

## Appendix S4: Alternative MARextinction

# Troubleshooting


