---
title: "MAR Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MAR Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```
## Overview

This vignette demonstrates the stepwise workflow for conducting a Mutation-Area
Relationship (MAR) analysis using the `mar` package, illustrated with the included
 *Arabidopsis thaliana* demo data.

```{r setup}
library(mar)
```
## Step 1: Import Raw Data

Begin by locating and importing the bundled example data included with the
`mar` package. `system.file()` resolves the path to files stored in
`inst/extdata/`, ensuring the vignette works regardless of where the package
is installed. `lonlat_parser()` reads the coordinate file (TXT, CSV, or TSV) and 
returns a data frame of sample IDs with their longitude and latitude, while
`text_parser()` reads the genotype matrix and returns a `margeno` object
containing the genotype data alongside sample and variant metadata.

```{r import}
# Sample coordinates
lonlatfile <- system.file("extdata", "lonlat.csv", package = "mar")

# Raw genotype matrix
genofile <- system.file("extdata", "genome.tsv", package = "mar")

# Then pass those paths to the parsers
coords <- lonlat_parser(lonlatfile)
geno_raw <- text_parser(genofile)

# inspect file format
head(coords)
print(geno_raw)
```

## Step 2: Prepare Spatial and Genotype Objects

### Spatial map object

`marmaps()` projects coordinates onto a raster grid at the chosen resolution
and coordinate reference system.

```{r spatial}
sample_maps <- marmaps(
  coords,
  mapres = 0.1,
  mapcrs = "EPSG:4326"
)
```

### Genotype object

`margeno()` packages the raw genotype matrix together with required metadata.
The `sample.id` vector **must** match the identifiers in the spatial object.

```{r genotype}
geno_obj <- margeno(
  genotype   = geno_raw$genotype,
  sample.id  = sample_maps$sample.id,
  variant.id = seq_len(nrow(geno_raw$genotype)),
  position   = rep(1:10, 1000),
  chromosome = sample(1:5, 10000, replace = TRUE),
  ploidy     = 2
)
```

### Combined genomic-spatial object

`genomaps()` merges the genotype and spatial objects into a single analysis
object used by all downstream functions.

```{r combine}
main_geno <- genomaps(geno_obj, sample_maps)
```

## Step 3: Explore the Site Frequency Spectrum (SFS)

The observed SFS summarises how allele counts are distributed across sites;
the expected SFS provides a neutral-theory baseline for comparison.

```{r sfs}
# Extract allele count vector for one locus
AC <- geno_obj$genotype[1, ]
N <- ncol(geno_obj$genotype)
ploidy <- geno_obj$ploidy

# Observed SFS
obs_sfs <- sfs(AC, N, ploidy = ploidy)

# Expected (neutral) SFS for a sample of 1004 diploid individuals
exp_sfs <- expsfs(500, 50, ploidy = 2)

# Visual comparison
plot(obs_sfs, exp_sfs)
```

## Step 4: Estimate the MAR

The MAR is estimated in four sub-steps.

### 4a. Generate the diversity-by-area sampling table

```{r mar-sampling}
# uses of myseed not neccessary, used here for reproducability
mar_sampling <- MARsampling(main_geno, myseed = 91)
```

### 4b. Calculate grid-cell areas from the raster map

```{r area}
gm_area <- mar:::.areaofraster(main_geno$maps$samplemap)
```

### 4c. Calculate gridded mutation / genetic diversity

`mutdiv.gridded()` computes per-cell diversity across the landscape within the
supplied bounding box (xmin, xmax, ymin, ymax in the map CRS).

```{r mutdiv}
mut_div <- mutdiv.gridded(
  main_geno,
  gmarea = gm_area,
  bbox   = c(10, 70, -10, 120)
)
```

### 4d. Fit the mutation-area relationship model

```{r mar-fit}
mar_fit <- MARcalc(mar_sampling)
marlist <- mar:::.pipe_MARcalc(mar_sampling)
```

## Step 5: Estimate Extinction / Diversity Loss

`MARextinction()` projects the expected loss of genetic diversity under
hypothetical area reduction scenarios.

```{r mar-extinction}
mar_ext <- MARextinction(main_geno)
plot(mar_ext)
```

## Step 7: Inspect and Export Outputs

Use `str()`, `print()`, and `summary()` to inspect any result object.

```{r inspect}
# Object structures
str(mar_sampling)
str(mut_div)
str(mar_fit)
str(mar_ext)

# Model summaries
print(mar_fit)
summary(mar_fit)
```

To persist results, use standard R serialisation or export to CSV:

```{r export}
saveRDS(mar_fit, file = "mar_fit.rds")
write.csv(as.data.frame(mar_sampling), file = "mar_sampling.csv", row.names = FALSE)
```
